Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-04-18 manuscript.qmd
add manuscript qmd

---
title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - HPO
  - chatGPT
  - rare disease 
bibliography: bibliography.bib
---

@@

p1 <- ggplot2::ggplot(data = dat_top,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = element_text(size=16)) + 
  labs(x = "Annotation",
       y = "HPO term", 
       fill = "")

# Heatmap of severity scores for top phenotypes
p2 <- ggplot2::ggplot(data = dat_middle,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = ggplot2::element_text(size=16)) + 
  ggplot2::labs(x = "Annotation",
                y = "HPO term", 
                fill = "")

# Heatmap of severity scores for top phenotypes
p3 <- ggplot2::ggplot(data = dat_bottom,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_ghibli_d("YesterdayMedium", direction = -1) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right', 
                 text = ggplot2::element_text(size=16)) + 
  ggplot2::labs(x = "Annotation",
                y = "HPO term", 
                fill = "")

# Combine the plots
p_combined <- patchwork::wrap_plots(p1, p2, p3, ncol = 3)

# Print the combined plot
print(p_combined)
```

>>>>

---
title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - HPO
  - chatGPT
  - rare disease 
bibliography: references.bib
---

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(HPOExplorer)
```

@@

p1 <- ggplot2::ggplot(data = dat_all,
                      ggplot2::aes(x=variable, y=hpo_name, fill=value)) +
  ggplot2::geom_tile(colour = "white", lwd = 0.5, linetype =1) +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_brewer(palette = "GnBu", direction = -1) +
  ggplot2::theme_classic() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.position = 'right',
                 text = element_text(size=16)) +
  labs(x = "Clinical characteristic",
       y = "HPO term",
       fill = "") +
  ggplot2::facet_grid(position ~ ., scales = "free_y")

p2 <- ggplot2::ggplot(data = dat_all,
                      ggplot2::aes(x="severity_score_gpt", y=hpo_name,
                                   fill=severity_score_gpt)) +
  ggplot2::geom_tile() +
  ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_gradientn(colours = colorspace::heat_hcl(7), trans = "reverse") +
  ggplot2::labs(x=NULL, y=NULL, fill = "Severity score") +
  ggplot2::theme_void() +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                 axis.ticks.y = ggplot2::element_blank(),
                 legend.position = 'right') +
  guides(fill = guide_colorbar(reverse = TRUE))

p3 <- patchwork::wrap_plots(p1, p2, ncol = 2,
                            widths = c(1,.2),
                            guides = "collect")

print(p3)
